\chapter{Algebraic Effect Handlers}
\label{chap:effects}

Algebraic Effects is the mathematical model for computational effects first introduced in \cite{plotkin2003algebraic} as an extension to the $\lambda$-calculus. After some time, Effect Handlers are introduced in \cite{plotkin2013handling} as a new addition to programming languages. Algebraic Effects and Handlers allows programmers to write programs with complex control flows. Functionalities such as exception handling, asynchronousity, I/O operations, and others are proven to be implement-able using Algebraic Effect Handlers. In this thesis, we use the term Algebraic Effect Handlers, Effect Handlers, and Handlers interchangably.

Algebraic Effect Handlers adapts the use of Call-By-Push-Value \cite{levy2013call} to separate value types, and computation types. This separation helps typing the computation together with possible effects, whereas value types are pure. This formal model usually provides reduction rules for its semantics.

In another spectrum, Effect Handlers are modeled by a combination of row types on top of \systemfw, named \systemfe \cite{xie2020effect}. The type rules are adjusted to accompany with a list of effects for an expression. This formal model usually provides evaluation contexts for its semantics.

\section{Examples}

In this section, we introduce some examples to breakdown the semantics of Effect Handlers. First we introduce how exceptions can be encoded using effect handlers, then we show how state management can also be encoded using effect handlers. First we define the exception effect:

$\attrE{\Sigma}(exn) = \{ throw : \eTForall{\alpha}{string\to\alpha} \}$

$throw-on-zero = \eLam{x:num}{\eIf{zero?(x)}{\ePerform{throw}{num}\ \text{``must not be zero''}}{x}}$

$h = \{throw \to \attrE{\Lambda \alpha^k.\eLam{msg:string}{\eLam{k}{42}}} \}$

$main = (\eHandle{h}{throw-on-zero\ 0} + 1) + 8$

The program evaluates to $50$ because the argument provided $0$ trigger the $throw$ handler. This handler will \textbf{ignore the current computation}, and return a default value $42$.

Readers familiar with try/catch (Java, Javascript) or try/except (Python) can try to think the handler as beginning a try scope; with $throw$ defined in $h$ as the catch/except scope. When a throw is called, the rest of the current computation is ignored. Using effect handlers, that is equivalent to ignore parameter $k$ in $throw$ and return a default value. Then the code continues outside the handle with the default value and adds $8$. The equivalent code in Python would be in Listing \ref{code:effect-python}.

\begin{listing}
\begin{minted}{python}
def throw_on_zero(x):
  if x == 0:
    raise Throw("must not be zero")
  return x

def handle_throw_on_zero():
  try:
    return throw_on_zero(0) + 1
  except Throw(msg):
    return 42

def main():
    handle_throw_on_zero() + 8
\end{minted}
\caption{Python equivalent example of effect handlers}
\label{code:effects1}
\end{listing}


\section{Language \lange}

In this section, we introduce \lange, a formal model of the Algebraic Effect Handlers as a model language for our target language for the compiler. Previously mentioned, we have two main choices for the formal language, (i) in the flavour of fine-grain call-by-value or (ii) in the flavour of \systemfe. We have chosen to adapt the (ii) flavour into the formal language due to its evaluation context semantics, which is helpful for our proofs.

We adopt and extend the language of \systemfe\ such that it can support all expressions available to \langc, including number type, binary operations, and if expression. We also introduce a new expression \eBlame{p} blaming the party \attrE{p}.

The language is polymorphic, with higher-kinded type system where effect is treated as a different kind from basic type, where in traditional effect type system, they are expressed in computational type. Effects are represented using row types, each row is indexed by a label. The environment of effect labels $\attrE{\Sigma}$ maps each labels to operations.

Programmers can define operations to be handled by rows of \attrE{h}, consisting of operation names, and their behaviors. A handler is also a value, and can be applied with a function ($\eHandler{h} \ \attrE{f}$) which a shorthand to a handle expression with $\attrE{f}$ applied to unit. In general, $\eHandle{h}{e}$ is used to run $\attrE{e}$ with effect handler $\attrE{h}$. During evaluation of $\attrE{e}$, an operation can be invoked by applying $\ePerform{op}{\sigma}$ to a value. If $\attrE{op}$ is declared in any handler of the current context, then the operation is handled using the closest handler. Because we define \lange\ well-typedness as no unhandled effects, there is no such case that an operation is undeclared once the program is well-typed.

A basic value, booleans and numbers, has no effect. A function $\attrE{\sigma_1\to\epsilon\sigma_2}$ receiving a type $\attrE{\sigma_1}$ producing a $\attrE{\sigma_2}$ value can invoke effects in $\attrE{\epsilon}$. A handler $\attrE{h} = \attrE{\{} \attrE{op_1 \to f_1}, \attrE{\dots}, \attrE{op_n \to f_n} \attrE{\}}$ must unify all operation types into one, allowing effect label of these operations, and other effects allowed by the environment. Because there is no type unification, therefore, these operations must return the same type.

Type judgement $\eTypesC{\eGamma}{e}{\sigma}{\epsilon}$ keeps a list of effects $\attrE{\epsilon}$ available to the expression $\attrE{e}$ being typed with type $\attrE{\sigma}$.

Following \cite{xie2020effect}, we adopt the \textit{dot notation} to decompose and compose evaluation contexts, as defined in Figure \ref{fig:lange-dot}. We would write $v \cdot \text{handle } h \cdot E \cdot e$ as a shorthand for $v \, (\text{handle } h \, (E[e]))$.

This thesis introduces two languages. To effectively distinguish the languages used, we have chosen different typeset styling for both languages. \lange\ will be typeset in \attrE{red\ serif}. The reduction arrow of \lange\ is $\eStep$.

\begin{singlespace}

% ==========================================
% DOT NOTATION FOR LANGUAGE E
% ==========================================
\begin{figure}[h]
  \centering
  \begin{tabular}{llll}
    Context Composition & $\attrE{E} \eDot \attrE{e}$ & $\doteq$ & $\attrE{E[e]}$ \\
    Argument Evaluation & $\attrE{\eHole \ e} \eDot \attrE{E}$ & $\doteq$ & $\attrE{E \ e}$ \\
    Function Evaluation & $\attrE{v \ \eHole} \eDot \attrE{E}$ & $\doteq$ & $\attrE{v} \eDot \attrE{E} \ \doteq \ \attrE{v \ E}$ \\
    Handler Application & $\eHandle{h}{\eHole} \eDot \attrE{E}$ & $\doteq$ & $\eHandle{h}{} \eDot \attrE{E} \ \doteq \eHandle{h}{E}$
  \end{tabular}
  \caption{$\lange$ Evaluation Context Dot Notation}
  \label{fig:lange-dot}
\end{figure}

% ==========================================
% SYNTAX FOR LANGUAGE E
% ==========================================
\begin{figure}[h]
  \centering
  \begin{tabular}{llll}
    Types       & $\attrE{\sigma}$ & $::=$ & $\attrE{\alpha^k} \mid \attrE{c^k \ \sigma \dots \sigma} \mid \attrE{\sigma \to \epsilon\sigma} \mid \attrE{\forall \alpha^k. \sigma}$ \\
    Kinds       & $\attrE{k}$      & $::=$ & $\attrE{*} \mid \attrE{k \to k} \mid \eKw{eff} \mid \eKw{lab}$ \\
    \\
    Expressions & $\attrE{e}$      & $::=$ & $\attrE{v} \mid \attrE{e_1 \ e_2} \mid \attrE{e[\sigma]} \mid \eHandle{h}{e}$ \\
    & & $\mid$ & $\attrE{e_1 + e_2} \mid \attrE{e_1 - e_2} \mid \attrE{e_1 \wedge e_2} \mid \attrE{e_1 \vee e_2}$ \\
    & & $\mid$ & $\eIf{e_1}{e_2}{e_3} \mid \attrE{zero?(e)}$ \\
    & & $\mid$ & $\eBlame{p}$\\

    Values      & $\attrE{v}$      & $::=$ & $\attrE{x} \mid \attrE{0}, \attrE{1}, ..., \attrE{N} \mid \eKw{tt} \mid \eKw{ff} \mid \eLam{x:\sigma}{e}$ \\
    & & $\mid$ & $\attrE{\Lambda \alpha^k. v} \mid \eHandler{h} \mid \ePerform{op}{\overline{\sigma}}$ \\
    Handlers    & $\attrE{h}$      & $::=$ & $\attrE{\{} \attrE{op_1 \to f_1}, \attrE{\dots}, \attrE{op_n \to f_n} \attrE{\}}$ \\
    Eval Contexts & $\attrE{F}$    & $::=$ & $\eHole \mid \attrE{F \ e} \mid \attrE{v \ F} \mid \attrE{F[\sigma]}$ \\
                & $\attrE{E}$      & $::=$ & $\eHole \mid \attrE{E \ e} \mid \attrE{v \ E} \mid \attrE{E[\sigma]} \mid \eHandle{h}{E}$ \\
  \end{tabular}
  \caption{Syntax of Language $\lange$}
  \label{fig:lange-syntax}
\end{figure}

% ==========================================
% TYPE RULES FOR LANGUAGE E
% ==========================================
\begin{figure}[htbp]
  \centering
  \begin{mathpar}
  \inferrule[Type-E-X-Val]
    {\attrE{x : \sigma} \in \eGamma}
    {\eTypesV{\eGamma}{x}{\sigma}}

  \inferrule[Type-E-Lambda]
    {\eTypesC{\eGamma, \attrE{x : \sigma_1}}{e}{\sigma_2}{\epsilon}}
    {\eTypesV{\eGamma}{\eLam{x:\sigma_1}{e}}{\eTArr{\sigma_1}{\epsilon\sigma_2}}}

  \inferrule[Type-E-Val]
    {\eTypesV{\eGamma}{v}{\sigma}}
    {\eTypesC{\eGamma}{v}{\sigma}{\epsilon}}

  \inferrule[Type-E-TypeLambda]
    {\eTypesV{\eGamma}{v}{\sigma} \\ \attrE{k} \neq \eKw{lab}}
    {\eTypesV{\eGamma}{\eTAbs{\alpha^k}{v}}{\eTForall{\alpha^k}{\sigma}}}

  \inferrule[Type-E-If]
    { \eTypesC{\eGamma}{e_1}{bool}{\epsilon} \\
      \eTypesC{\eGamma}{e_2}{\sigma}{\epsilon} \\
      \eTypesC{\eGamma}{e_3}{\sigma}{\epsilon} }
    { \eTypesC{\eGamma}{\eIf{e_1}{e_2}{e_3}}{\sigma}{\epsilon} }

  \inferrule[Type-E-App]
    {\eTypesC{\eGamma}{e_1}{\eTArr{\sigma_1}{\sigma}}{\epsilon} \\
     \eTypesC{\eGamma}{e_2}{\sigma_1}{\epsilon}}
    {\eTypesC{\eGamma}{e_1 \, e_2}{\sigma}{\epsilon}}

  \inferrule[Type-E-TypeApp]
    {\eTypesC{\eGamma}{e}{\eTForall{\alpha^k}{\sigma_1}}{\epsilon} \\
     \eWf{\sigma}{k}}
    {\eTypesC{\eGamma}{e[\sigma]}{\sigma_1[\alpha := \sigma]}{\epsilon}}

  \inferrule[Type-E-Perform]
    {\attrE{op : \forall \overline{\alpha} . \sigma_1 \to \sigma_2} \in \attrE{\Sigma}(\attrE{l}) \\
     \attrE{\overline{\alpha}} \notin \text{ftv}(\eGamma)}
    {\eTypesV{\eGamma}{\ePerform{op}{\overline{\sigma}}}{\eTArr{\sigma_1[\overline{\alpha} := \overline{\sigma}]}{\eEff{l}{\epsilon} \, \sigma_2[\overline{\alpha} := \overline{\sigma}]}}}

  \inferrule[Type-E-Ops]
    {\attrE{op_i : \forall \overline{\alpha} . \sigma_1 \to \sigma_2} \in \attrE{\Sigma}(\attrE{l}) \\
     \attrE{\overline{\alpha}} \notin \text{ftv}(\attrE{\epsilon \, \sigma}) \\
     \eTypesV{\eGamma}{f_i}{\eTForall{\overline{\alpha}}{\eTArr{\sigma_1}{\eTArr{(\eTArr{\sigma_2}{\epsilon\sigma})}{\epsilon\sigma}}}}}
    {\eTypesOps{\eGamma}{\{ op_1 \to f_1, \dots, op_n \to f_n \}}{\sigma}{l}{\epsilon}}

  \inferrule[Type-E-Handler]
    {\eTypesOps{\eGamma}{h}{\sigma}{l}{\epsilon}}
    {\eTypesV{\eGamma}{\eHandler{h}}{\eTArr{(\eTArr{()}{\eEff{l}{\epsilon} \, \sigma})}{\sigma}}}

  \inferrule[Type-E-Handle]
    {\eTypesOps{\eGamma}{h}{\sigma}{l}{\epsilon} \\
     \eTypesC{\eGamma}{e}{\sigma}{\eEff{l}{\epsilon}}}
    {\eTypesC{\eGamma}{\eHandle{h}{e}}{\sigma}{\epsilon}}
  \end{mathpar}
  \caption{Type Rules for System $\lange$}
  \label{fig:lange-typing}
\end{figure}

% ==========================================
% REDUCTION RULES FOR LANGUAGE E
% ==========================================
\begin{figure}[htbp]
  \centering
  \begin{mathpar}
  \inferrule[Step-E-App]
    {\strut}
    {(\eLam{x:\sigma}{e}) \ \attrE{v} \eStep \attrE{e[x := v]}}

  \inferrule[Step-E-TApp]
    {\strut}
    {(\eTAbs{\alpha^k}{v}) \attrE{[\sigma]} \eStep \attrE{v[\alpha := \sigma]}}

  \inferrule[Step-E-Handler]
    {\strut}
    {(\eHandler{h}) \ \attrE{v} \eStep \eKw{handle} \ \attrE{h} \eDot \attrE{v \ ()}}

  \inferrule[Step-E-Return]
    {\strut}
    {\eKw{handle} \ \attrE{h} \eDot \attrE{v} \eStep \attrE{v}}

  \inferrule[Step-E-Perform]
    {\attrE{op} \notin \mathrm{bop}(\attrE{E}) \wedge (\attrE{op \to f}) \in \attrE{h} \\
     \attrE{op : \forall \overline{\alpha} . \sigma_1 \to \sigma_2} \in \attrE{\Sigma}(\attrE{l}) \\
     \attrE{k} = \eLam{x : \sigma_2[\overline{\alpha} := \overline{\sigma}]}{\eKw{handle} \ \attrE{h} \eDot \attrE{E} \eDot \attrE{x}}}
    {\eKw{handle} \ \attrE{h} \eDot \attrE{E} \eDot \ePerform{op}{\overline{\sigma}} \ \attrE{v} \eStep \attrE{f[\overline{\sigma}] \ v \ k}}

  \inferrule[Step-E-Context]
    { \attrE{e_1} \eStep \attrE{e_2} }
    { \attrE{E[e_1]} \eStep \attrE{E[e_2]} }

  \inferrule[Step-E-Blame]
    { \attrE{E} \neq \eHole }
    { \attrE{E[blame \ k]} \eStep \attrE{blame \ k} }

  \inferrule[Step-E-Refl]
    { \strut }
    { \attrE{e} \eStepMany \attrE{e} }

  \inferrule[Step-E-Trans]
    { \attrE{e} \eStep \attrE{e_1} \\
      \attrE{e_1} \eStepMany \attrE{e_2} }
    { \attrE{e} \eStepMany \attrE{e_2} }

  \end{mathpar}
  \caption{Reduction Rules for System $\lange$}
  \label{fig:lange-reduction}
\end{figure}
\end{singlespace}

\section{Soundness and Correctness}

Our algebraic effects handler system has type soundness by progress () and preservation ().


\begin{theorem}[Progress for \lange]
  \label{theorem:lange-progress}
  If $\eTypesC{\emptyset}{e_1}{\sigma}{\langle \rangle}$, then either $\attrE{e_1}$ is a value, or there exists an $\attrE{e_2}$ such that $\attrE{e_1} \eStep \attrE{e_2}$.
\end{theorem}

\begin{theorem}[Preservation for \lange]
  \label{theorem:lange-preservation}
  If $\eTypesC{\emptyset}{e_1}{\sigma}{\langle \rangle}$ and $\attrE{e_1} \eStep \attrE{e_2}$, then $\eTypesC{\emptyset}{e_2}{\sigma}{\langle \rangle}$.
\end{theorem}

Our effect system is also safe, or all effects are handled or specified in the judgement. To express safety of effects, we first define (in Figure \ref{fig:lange-metafunc}), $\lceil \attrE{E} \rceil^l$ extracts all labels in the context; and $\text{bop}(\attrE{E})$ extracts all $\attrE{op}$ in the context.

\begin{figure}
\[
  \lceil \attrE{E} \rceil^l = \lceil F_0 \cdot \text{handle } h_1 \cdot F_1 \cdot \dots \cdot \text{handle } h_n \cdot F_n \rceil^l = \langle l_n, \dots, l_1 \rangle \text{ iff } h_i : \Sigma(l_i)
\]

\begin{align*}
  \text{bop}(\square) &= \emptyset \\
  \text{bop}(E \ e) &= \text{bop}(E) \\
  \text{bop}(v \ E) &= \text{bop}(E) \\
  \text{bop}(\text{handle} \ h \ E) &= \text{bop}(E) \cup \{ op \mid (op \to f) \in h \}
\end{align*}
\caption{Metafunction for context in \lange.}
\label{fig:lange-metafunc}
\end{figure}

\begin{lemma}[Well-typed operations are handled]
  \label{lemma:lange-op-handled}
If $\eTypesC{\emptyset}{E[\ePerform{op}{\overline{\sigma}} \ v]}{\sigma}{\langle \rangle}$, then $E$ has the form $E_1 \cdot \eHandler{h} \cdot E_2$ with $\attrE{op} \notin \text{bop}(E_2)$ and $\attrE{op \to f} \in \attrE{h}$.
\end{lemma}

\begin{lemma}[Effects types are meaningful]
  \label{lemma:lange-types-meaningful}
If $\eTypesC{\emptyset}{E[\ePerform{op}{\overline{\sigma}} \ v]}{\sigma}{\epsilon}$ with $\attrE{op} \notin \text{bop}(E)$, then $\attrE{op} \in \attrE{\Sigma}(l)$ and $l \in \attrE{\epsilon}$.
\end{lemma}
