\chapter{Introduction}

Software contracts, or \textit{behavioral contracts}, allow programmers to
attach assertions to values, and functions. These assertions are runtime
guarantees, on violation, an error is thrown, blaming the violating party.
Contracts are most useful when placed on a function, enforcing the function's
input and output values. While first-order functional contracts are useful,
they are too simple. Higher-order functional contracts allows programmers to
enforce contracts for higher-order functions. Dependent contracts are
functional contracts allowing usage of the function argument during
post-condition contract checking.

Recently, algebraic effect handlers has gained a large attraction, and is
adopted in OCaml since version 5.0. Effect handlers open a new way to identify
effects and handling them. Handlers are defined to handle one or many effects.
A handler for an effect can stop the program or continue with some result.
Handlers do not have state by default, but it can be implemented by applying
the new states to the continuation closure, and let the underlying computation
receives a state argument.

Combining the contract system and effect handlers is a new approach. Effectful
Contract \cite{10.1145/3632930} demonstrates how both systems can be combined,
providing contracts for effectful code. Contracts in their system remain a
language feature. In this dissertation, instead of combining the effects and
contracts, we ask ourselves whether contract checking itself can become an
effect, and reuse the capabilities of handlers.

\section{Contracts checking}

Software contracts, despite their usefulness, are not natively support in many
programming languages. Racket and D-lang are two languages that have native
support for contracts. C++, or its 2026 version, has some basic support for
contracts. Python has a proposal (PEP
316\footnote{\href{https://peps.python.org/pep-0316/}{https://peps.python.org/pep-0316/}}),
but it was deferred since 2003.

Racket implements contracts by code rewriting with macros. Moreover, protected
values and functions are wrapped inside a native proxy object (chaperone).
Native support allows Racket to implement higher-order contracts.

Without native support, the community have come up with strategies for implementing
software contracts. Java's libraries (C4J\footnote{\href{https://c4j-team.github.io/C4J/index.html}{https://c4j-team.github.io/C4J/index.html}}
and
Cofoja\footnote{\href{https://github.com/nhatminhle/cofoja}{https://github.com/nhatminhle/cofoja}}),
define contracts using Java's annotations and rewrite the bytecode during
compilation to add contract checking code. Similar strategies are used by Rust
(contracts\footnote{\href{https://crates.io/crates/contracts}{https://crates.io/crates/contracts}})
and Golang (gocontracts\footnote{\href{https://github.com/Parquery/gocontracts}{https://github.com/Parquery/gocontracts}})
libraries, where the code is rewritten to add pre and post condition code
checking to a function.

Python has support for first-class functions, and \textit{decorators}.
Decorators syntactically wraps a function using another function while
the original function name can still be referenced. This feature is used in many
Python libraries
(icontract\footnote{\href{https://github.com/Parquery/icontract}{https://github.com/Parquery/icontract}},
deal\footnote{\href{https://github.com/life4/deal}{https://github.com/life4/deal}},
etc.) to implement functional contracts.

Without native contract support, functional and dependent contracts are
feasible. While higher-order contracts remains a challenge to implement.

\section{Contracts checking as effects}

Contracts checking adds extra computation, when the checks pass, the program
continues. When the checks fail, the program aborts. The behavior is similar to
throwing an exception. Exception is one kind of side-effect. The similarities
between exceptions and contracts checking are sufficient to make contracts
checking an effect and use handlers to handle them.

Once contract effects are added, handlers need to be added. Correctly grouping
the handlers can share the context of contracts. Sharing the context between
contracts allows for optimization through contract caching, or cross-function
call data sharing. Previously, cross-function call data is stored in a global
storage \cite{MOY_FELLEISEN_2023}, or using effects to provide ``ghost'' state
\cite{10.1145/3632930}.

\section{Contributions}

% Algebraic Effects are sufficient to implement a basic Contract system on top of language.

% - Implementation in Ocaml
% - A use case of the contract system
% - A model as a blueprint to explain design
% - Basic properties of that model
% - Comparision with Racket contract system

This thesis explores the ``contracts checking as effects'' approach to
implementing software contracts. Through experiments, we show that this
approach accomodates functional contracts, dependent contracts, and the
challenging higher-order contracts. We also discovered that as effects, their
contexts can be shared if they are under the same handler. Sharing contracts
context opens new capabilities to optimizations and contract state sharing.
This disseration shows the above idea by providing a compiler model between
CPCF \cite{cpcf} and System $F^\epsilon$ \cite{xie2020effect}. We also provide
an implementation in OCaml as a practical implementation.

The rest of the thesis is organized as follows: Chapter \ref{chap:contracts}
and Chapter \ref{chap:effects} introduces the two formal languages \langc\
based on CPCF, and \lange\ based on System $F^\epsilon$; Chapter
\ref{chap:compiler} discusses a theoretical compiler from \langc\ to \lange\ as
a demonstration to ``contract checking as effects''; Chapter \ref{chap:ocaml}
introduces the OCaml language and its native support for effect handlers;
Chapter \ref{chap:implementation} provides the implementation of the
theoretical compiler in OCaml using PPX rewritter; Chapter \ref{chap:compare}
compares our implementation in OCaml to the Racket's contract system; Chapter
\ref{chap:final} concludes our remarks.
